---
title: "Homework 1"
author: "Siim Sepp"
date: "1/21/2025"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Loading packages
If your script depends on external packages, load them at the beginning. This shows users early on what the script dependencies are.  
```{r package ambiguity, message=TRUE}
ip <- installed.packages()
packages_used <- c("tidyverse", "workflowr")

for (package in packages_used){
  isInstalled <- package %in% rownames(ip)
  if (!isInstalled){
    stop(paste("ERROR: you need to install the", package, "package"))
  } else{
    library(package, character.only=T)
  }
}
```



#Create a loop that will sample variables from a normal distribution 100 times. **(1 pt)**
```{r}
for (i in 1:100) {
  sample<- rnorm(1)
}
```

- In each loop sample 10 values from a normal distribution with mean= 5 and variance =9. **(1 pt)** 

```{r}
for (i in 1:100) {
  sample<- rnorm(10, 5, sd=3)
}
```


- For each loop calculate and store the mean and variance of the sample values using vector multiplication. The means and variance estimates should be stored in separate vectors. **(1 pt)**.

```{r}
means<- NULL
vars<- NULL
for (i in 1:100) {
  sample<- rnorm(10, 5, sd=3)
  id<- rep(1, length(sample))
  mean<- sample %*% id / length(sample)
  means<- cbind(means, mean) # stores the means of the sample
  devs<- NULL
  for (p in sample) {
    dev<- (mean-p)^2
    devs<- cbind(devs, dev)
    
  }
  var<- devs %*% id /(length(sample) -1)
  vars<- cbind(vars, var) # stores the variances of the sample
}

```


- Visualize the distribution of the means a variance estimates using the hist() function. **(1 pt)**
```{r}
hist(means)
hist(vars)
```

Describe any differences in the distributions of the means and variance estimates. **(1 pt)**


The distribution of the variances seems to be more wider and centered around 9. 
The distribution of means seems to be more narrow and centered around 5. 


4) Put the markdown file in the appropriate folder and build the html files using workflowr. **(2.5 pts)**

 
