---
title: "HW5"
author: "sss322"
date: "2025-02-18"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Genetic Relationships: Identity by Descent
### The genetic (co)variance between individuals for a given phenotype is determined by common alleles for the QTL that explain variance for the phenotype

### In the absence of detailed information on the QTL, the (co)variance can be approximated by calculating the probability two individuals carry alleles that are identical by descent (IBD).
### - IBD represents the probability that the same allele is inherited by two individuals from some common ancestor.
### - These calculations can be done using pedigrees or genotypes

## A simple illustration
### In this exmple we are testing lines from 2 unrelated full sib families. The code below will simulate a polygenetic trait and compare prediction accuracy using a model that ignores genetic relationships and a mixed model that accounts for the relationships. ***you will need to download and install the package MASS***

##Setting up parameters

```{r paremeters}
library(MASS)
library(here)
here::i_am("analysis/Lab_5.Rmd")
#setting variety and residual variance
VarA=75
VarE=25
#mean of the families relative to the overall mean
meanFam=c(rep(-5,15),rep(5,15))
```

## Generating relationship matrix

```{r Relationship Matirx}
source(here::here("code/Simulate_and_Solve.R"))
# this is the expected relationship matrix for to unrelated full-sib families
# there is a total of 30 lines, 15 lines in each family
#initializing a 30 x 30 matrix with all values = 0
A=matrix(0,30,30)
#creating a block of .5 for family 1
A[1:15,1:15]=.5
#creating a block of .5 for family 2
A[16:30,16:30]=.5
#setting the diagonal to 1
diag(A)<-1
print("A matrix")
image(A)
```
## Simulate Breeding Values

```{r simulate and solve}
set.seed(35783948)
# sampling 30 variety effects
# now we are sampling from mvnorm since the variety effects are no longer iid
BV=MASS::mvrnorm(1,meanFam,VarA*A)
sol=simulate_and_solve(BV,A,VarA,VarE)
# LSM - Least squares means
# MM - BLUPs from the mixed model
```


## Questions 
 1)Conduct the following simulations using a difference in mean performance between the two families of 10.
  
   A.(1 pt)  Use VarA = 75 and VarE = 25 (heritability = .75). Change the seeds and run the simulation 5 times 
      - Calculate the correlations between LSM and the true BV
      - Calculate the correlations between MM and the true BV
      - Plot the BLUPs against LSM 
```{r}
set.seed(NULL)
VarA=75
VarE=25
meanFam=c(rep(-5,15),rep(5,15))
sols<- data.frame()
n=5

for (i in n){
   set.seed(i)
BV=MASS::mvrnorm(1,meanFam,VarA*A)
sol=simulate_and_solve(BV,A,VarA,VarE)
sol<- cbind(sol, BV)
sols<- rbind(sols, sol)

}
cor(sols$LSM, sols$BV)
cor(sols$MM, sols$BV)
plot(sols$MM, sols$LSM)
```
  
   B.(1 pt) Use VarA = 50 and VarE = 50 (heritability = .5). Change the seeds and run the simulation 5 times 
      - calculate the correlations between LSM and the true BV
      - calculate the correlations between MM and the true BV
      - Plot the BLUPs against LSM 
 
```{r}
set.seed(NULL)
VarA=50
VarE=50
meanFam=c(rep(-5,15),rep(5,15))
sols<- data.frame()

for (i in n){
   set.seed(i)
BV=MASS::mvrnorm(1,meanFam,VarA*A)
sol=simulate_and_solve(BV,A,VarA,VarE)
sol<- cbind(sol, BV)
sols<- rbind(sols, sol)

}
cor(sols$LSM, sols$BV)
cor(sols$MM, sols$BV)
plot(sols$MM, sols$LSM)
```
    
   C.(1 pt) Use VarA = 25 and VarE = 75 (heritability = .25). Change the seeds and run the simulation 5 times 
      - Calculate the correlations between LSM and the true BV
      - Calculate the correlations between MM and the true BV
      - Plot the BLUPs against LSM 
```{r}
set.seed(NULL)
VarA=25
VarE=75
meanFam=c(rep(-5,15),rep(5,15))
sols<- data.frame()

for (i in n){
   set.seed(i)
BV=MASS::mvrnorm(1,meanFam,VarA*A)
sol=simulate_and_solve(BV,A,VarA,VarE)
sol<- cbind(sol, BV)
sols<- rbind(sols, sol)

}
cor(sols$LSM, sols$BV)
cor(sols$MM, sols$BV)
plot(sols$MM, sols$LSM)
```
    
      
   D. (1 pt) What trends do you see in the plots of the BLUPS versus LSM? What would be causing this pattern?

The pedigree used in the MM to calculate the BLUPs is causing the families to segregate to different regression lines. 
This family info is also providing higher correlations to the MM with the BV compared to to the LSM. 

The MM solutions also seem to have less spread likely due to shrinkage.



   E. (1 pt) How does heritability impact the relative performance BLUPs and LSM in terms of accuracy (correlations to the true BV)? Why?

LSM loses correlation faster with the true BV when the heritability is decreasing. I think it is also caused by shrinking in the MM where more uncertainity is reduced by shrinking towards the mean. Wheareas when there is no nois then it matters less and the LSM is also pretty accurate. 


   
  2) Change the difference in mean performance between families to 5.
   
   A) (3 pts) Repeat the simulations in outlined in  (1A-1C) 

```{r}

set.seed(NULL)
VarA=75
VarE=25
meanFam=c(rep(-2.5,15),rep(2.5,15))
sols<- data.frame()

for (i in n){
   set.seed(i)
BV=MASS::mvrnorm(1,meanFam,VarA*A)
sol=simulate_and_solve(BV,A,VarA,VarE)
sol<- cbind(sol, BV)
sols<- rbind(sols, sol)
}
cor(sols$LSM, sols$BV)
cor(sols$MM, sols$BV)
plot(sols$MM, sols$LSM)
```

```{r}
set.seed(NULL)
VarA=50
VarE=50
meanFam=c(rep(-2.5,15),rep(2.5,15))
sols<- data.frame()

for (i in n){
   set.seed(i)
BV=MASS::mvrnorm(1,meanFam,VarA*A)
sol=simulate_and_solve(BV,A,VarA,VarE)
sol<- cbind(sol, BV)
sols<- rbind(sols, sol)

}
cor(sols$LSM, sols$BV)
cor(sols$MM, sols$BV)
plot(sols$MM, sols$LSM)
```

```{r}
set.seed(NULL)
VarA=25
VarE=75
meanFam=c(rep(-2.5,15),rep(2.5,15))
sols<- data.frame()

for (i in n){
   set.seed(i)
BV=MASS::mvrnorm(1,meanFam,VarA*A)
sol=simulate_and_solve(BV,A,VarA,VarE)
sol<- cbind(sol, BV)
sols<- rbind(sols, sol)
}
cor(sols$LSM, sols$BV)
cor(sols$MM, sols$BV)
plot(sols$MM, sols$LSM)
```
      
   B) (1 pt) How did the change in the difference in mean performance between families impact the relative performance of BLUPs and LSM? Why? 
   
  (1 point for turning in the assignment on time)

The correlation with true BV increased for both MM and LSM. 

The pedigree is less predictive in this case, so the MM performance closer to the LSM performance. 

 
