---
title: "HW2"
author: "sss322"
date: "2025-01-28"
output: html_document
---

## Homework
### For the homework assignment we will simulate data, contruct the corresponding design matrix, and use matrix algebra to estimate the treatment effects. The simulated dataset will have 18 obersevations from 3 treatments.

1) Allocate a vector $\mathbf{y}$ to store the 18 phenotypes and then simulate observations from each treatment group. (2 pts)
    a. Sample y1-y6 (given treatment 1) from a normal distribution with mean 3 and variance 1
    b. Sample y7-y12 (given treatment 2) from a normal distribution with mean 9 and variance 1
    c. Sample y13-y18 (given treatment 3) from a normal distribution with mean 14 and variance 1

```{r}
y1<- rnorm(6, 3, 1)
y2<- rnorm(6, 9, 1)
y3<- rnorm(6, 14,1)
Y<- (c(y1, y2, y3))
Y<- as.matrix(Y)
```


2) Set up the appropriate design matrix $\mathbf{X}$ for the simulated data $\mathbf{y}$. (2 pts) 
```{r}
X= matrix(0, 18, 3)
X[1:6,1] = c(rep(1, 6))
X[7:12,2] = c(rep(1, 6))
X[13:18,3] = c(rep(1, 6))
```


3) Calculate $\mathbf{X^t}\mathbf{y}$. What does each value in $\mathbf{X^t}\mathbf{y}$ represent? (2 pts)
```{r}
x_Ty<- t(X) %*% Y #  
x_Ty
```
The values represent the sum of phenotypes in the treatment



4) Calculate $\mathbf{X^t}\mathbf{X}$. What does each value in $\mathbf{X^t}\mathbf{X}$ represent? (2 pts)
```{r}
x_Tx<-  t(X) %*% X 
x_Tx
```
The values represent the number of samples per treatment



5) Using $\mathbf{X^t}\mathbf{X}$ and $\mathbf{X^t}\mathbf{y}$ calculate $\mathbf{\hat{\beta}}$ where $\hat{\beta_{1}} = \frac{\sum_{j=1}^{6} y_{1j}}{6}$, $\hat{\beta_{2}} = \frac{\sum_{j=1}^{6} y_{2j}}{6}$, and $\hat{\beta_{3}} = \frac{\sum_{j=1}^{6} y_{3j}}{6}$. (2 pts)


```{r}
solve(x_Tx) %*% x_Ty
```





