---
title: "Problem Set 1"
author: "sss322"
date: "2025-02-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



#### **Q1 As a plant breeder I want to fix an allele that provides resistance to an important crop disease.** 
Unfortunately, I do not have the ability to directly genotype the causative mutation, but I have identified a genetic marker that is closely linked the causative mutation. Based on mapping experiments the recombination frequency between the marker (alleles: A & a) and the QTL (alleles: Q – resistant & q – susceptible) is r = .25I have crossed a homozygous resistant line to a homozygous susceptible line to get an offspringwith the genotype: !AQ$ %aq
(I will self this individual to generate offspring segregating for resistance.

#### **A. If I select only lines with the genotype AA, what expected proportion of the selected lines that will be homozygous for the favorable QTL (QQ)? (2 pts)**



In the whole genotype: 

Probability of AA is 0.25
(from the Punnett square)

Probability of AQ = P(no crossover) * 0.5 = (1 - 0.25) * 0.5 = 0.375


I want the find the probability of QQ given AA, through using the joint probability. 

P= P(QQ and AA) / P(AA) = (P(AQ) * P(AQ)) / P(AA)

P=  (0.375 * 0.375)) / 0.25 = 0.5625


The expected proportion of QQ given AA is equal to 0.5625


**The probability that i get the genotype where both marker alleles are A and there has been no recombination between the marker and the qtl is 0.5625**



#### **Q2 Similar to Q1 I have identified markers linked to a QTL of interest and I now want to practice marker assisted selection to fix the favorable allele in a population. ** 

In this scenario I have 2 markers linked on either side of the QTL. Marker 1 (alleles A, a) has a recombination frequency r =.4 with the QTL and Marker 2 (alleles B, b) has a recombination frequency r =.3 with the QTL.
As before I’ve crossed a line that is homozygous for the beneficial allele (Q) and a line that is homozygous for the unfavorable allele (q): )AQB+,aqb.
I selfed the heterozygous line above to generate many offspring.The QQ genotype has a value of a = 5 for the trait of interest, the unfavorable genotype qq has
a value of -a = -5 for the trait of interest. The trait is additive (no dominance)

#### **A. As a breeder I want to fix the resistance allele. If I select only lines with the genotype AABB, what is the expected proportion of the selected lines that will be homozygous for the favorable QTL (QQ)? (2 pts)**

P(QQ given AABB) = P(QQ and AABB) / P(AABB)

**Joint probability:**

P(AABB) =  P(no crossover or 2 crossovers in 1st haplotype) * P(no crossover or 2 crossovers in 2nd haplotype)

P(AABB)= (0.5 * 0.6 * 0.7 + 0.5 * 0.4 * 0.3) * (0.5 * 0.6 * 0.7 + 0.5 * 0.4 * 0.3) = 0.27*0.27= 0.0729

**Probability of QQ and AABB:**

Probability of QQ and AABB = 0.5 * (0.6 * 0.7) * 0.5 * (0.6 * 0.7) = 0.0441

**P(QQ given AABB) **

Probability of (QQ given AABB) = 0.0441/0.0729 = 0.6049383

**The proportion of lines that have QQ given AABB genotype is 0.61**



#### **B. What is the expected value for offspring with the genotype AABB? (3 pts)**
P(QQ given AABB) =  0.7778 ^2 =0.605.


P(qq given AABB)=  (P(2 crossovers have happened and AB)/P(AB))^2  =  (0.5 * 0.4 * 0.3) / (0.5 * 0.4 * 0.3 + 0.5 * 0.6 * 0.7)^2   =0.222 ^ 2 = 0.049

P(Qq given AABB) =  2 * 0.7778 * 0.2222 = 0.345 



```{r}
(0.605*5) + (0.345*0) + (0.049*-5) 
```
The expected value when the genotype is AABB is 2.7


#### **C. What is the expected value for lines with the genotype AA? (3 pts)**

Probability of QQ given AA = P(Q given A) * P(Q given A) = (P(Q and A)/P(A))^2 = ((0.5 * (1 - 0.4)) / 0.5)^2 = (0.5 * 0.6 / 0.5)^2  = 0.36

Probability of qq given AA = P(q given A) * P(q given A) = (P(Q and A)/P(A))^2 = ((0.5 * 0.4 / 0.5)^2 =  (0.4)^2 = 0.15

Probability of Qq given AA is 2 * 0.4 * 0.6 = 0.48


```{r}
(0.36*5) + (0.48*0) + (0.16*-5) 
```
Expected value whenthe genotype is AA =1 


#### **D. What is the expected value for lines with the genotype BB? (3 pts)**


Probability of QQ given BB = P(Q given B) * P(Q given B) = (P(Q and B)/P(B))^2 = ((0.5 * (1 - 0.3)) / 0.5)^2 = (0.5 * 0.7 / 0.5)^2  = 0.49

Probability of qq given BB = P(q given B) * P(q given B) = (P(q and B)/P(B))^2 = ((0.5 * 0.3 / 0.5)^2 = (0.5 * 0.3 / 0.5)^2  = 0.09

Probability of Qq given BB is 2 * 0.7 * 0.3 = 0.42


```{r}
(0.49*5) + (0.42*0) + (0.09*-5) 
```
Expected value when the genotype is BB =2 



#### **E. How close would a single marker need to be in order for marker assisted selection to as effective and the two marker example above? (2 pts)**

x= the 1-r value for the MAS to be as effective as two marker selection

q= 1-x

5x^2 + 2x(1-x) * 0 + (1-x)^2 * (-5) = 2.78

5x^2 + 0 + -5+10x-5p^2 = 2.7

10x = 2.7+5 = 7.78

x= 0.78

r= 1-x= 0.22

**The recombination frequency of a marker has to be smaller than 0.22 to be as effective as two marker selection in MAS. **




